Locality and Effect Reflection

### 2025-07-10 paper.tex
Fix up code examples

  module State_eff = Reflection(State)

  let handle_state (f : Handler(State).t @ local -> 'a) (i : int) : 'a =
    let rec loop s : 'a Term(State).t @ local -> 'a  = function
      | Return x -> x
      | Op(Get, k) -> loop s (k s)
      | Op(Set s', k) -> loop s' (k ())
    in
    loop i (State_eff.reify_local f)
\end{lstlisting}
which calls \lstinline[style=oxcaml]{f} and passes it a local value of
type \lstinline[style=oxcaml]{Handler(State).t}. For the duration of the
call to \lstinline[style=oxcaml]{f} the ambient monad is extended with
two additional effectful operations, which can be used via reflection:
\begin{lstlisting}[style=oxcaml]
  let get (h : Handler(State).t @ local) : int =
    State_eff.perform Get h

@@

Locality tracks whether a value can 'escape' the current
\emph{region}. By default, the current region ends when a function
returns, so local values cannot be returned from functions.
\begin{lstlisting}[style=oxcaml]
  let leak_local x =
    let (o @ local) = Some x in
    o
  (*@ \textcolor{red}{Error: o} @*)
         (*@ \textcolor{red}{\textasciicircum} @*)
  (*@ \textcolor{red}{This value is local and would escape its region.} @*)
\end{lstlisting}

@@

As shown in the introduction we can implement an effect handler for
state using this API. We start by defining the operations of the effect:
\begin{lstlisting}[style=oxcaml]
  module State = struct
    type 'a t =
      | Get : int t
      | Set : int -> unit t
  end
\end{lstlisting}
Then we instantiate the reflection isomorphism for that effect and use
it to write a handler:
\begin{lstlisting}[style=oxcaml]
  module State_eff = Reflection(State)

  let handle_state f i =
    let rec loop s = function
      | Term(State).Return x -> x
      | Term(State).Op(Get, k) -> loop s (k s)
      | Term(State).Op(Set s', k) -> loop s' (k ())
    in
    loop i (State_eff.reify_local f)
\end{lstlisting}
Finally we add an alias for the handler type and some convenient
wrappers for the generic effects:
\begin{lstlisting}[style=oxcaml]
  type handler = Handler(State).t

@@

  val handle_state : (handler @ local -> 'a) @ local -> int -> 'a
  val get : handler @ local -> int
  val set : handler @ local -> int -> unit
\end{lstlisting}
Note that we are able to use \lstinline[style=oxcaml]{reify_local} to allow
the function passed to \lstinline[style=oxcaml]{handle_state} to be local.

@@

  let handle_state f i = f (ref i)
  let get r = !r
  let set r i = r := i
\end{lstlisting}
This implementation gives the same behaviour as the one built using
effect reflection. Crucially it is just as easy to reason about, which
in turn means it is just as easy to reason about as the state monad. In
some sense, the improved reasoning ability of the state monad over
ordinary mutable state is precisely captured by the notion of locality.

@@

The following function uses state to calculate the total of a list of
integers:
\begin{lstlisting}[style=oxcaml]
  let total l =
    handle_state (fun h ->
      List.iter (fun x -> set h (get h + x)) l;
      get h)
\end{lstlisting}
This relies on the locality of the function parameter of
\lstinline[style=oxcaml]{List.iter}:
\begin{lstlisting}[style=oxcaml]
  val iter : ('a -> unit) @ local -> 'a list -> unit
\end{lstlisting}
The \lstinline[style=oxcaml]{local} annotation on the parameter ensures
that any function passed will not escape the current region. This allows
us to pass in a closure that closes over the
\lstinline[style=oxcaml]{local} handler \lstinline[style=oxcaml]{h} and
use it to perform some effects.

@@

A more realistic use of effect reflection is a \emph{generator}
effect. This effect has a single operation:
\begin{lstlisting}[style=oxcaml]
  module Gen = struct
    type 'a t =
      | Gen : int -> unit t
  end
\end{lstlisting}
Along with a handler that turns computations using
\lstinline[style=oxcaml]{Gen} into streams of integers.
\begin{lstlisting}[style=oxcaml]
  module Gen_eff = Reflection(Gen)

@@

  let handle_gen f =
    let rec loop = function
      | Term(Gen).Return () -> Finished
      | Term(Gen).Op(Gen i, k) -> More(i, k)
    in
    loop (Gen_eff.reify f)
\end{lstlisting}

@@

\begin{lstlisting}[style=oxcaml]
  module Await_eff = Reflection(Await)

  let handle_await f =
    let rec loop = function
      | Term(Await).Return x -> Deferred.return x
      | Term(Await).Op(Await d, k) ->
          Deferred.bind d (fun x -> loop (k x))
    in
    loop (Await_eff.reify f)

@@

A key feature of algebraic effects is how easy they are to compose. We
can use state whilst using asynchronous I/O by simply nesting their
handlers:
\begin{lstlisting}[style=oxcaml]
  handle_await (fun ch ->
    handle_state i (fun sh ->
      List.iter files (fun fl ->
        let s = await ch (File.read fl) in
        let i = int_of_string s in
        set sh (i + get sh));
      get sh))
\end{lstlisting}

@@

If we tried to do something similar using our
\lstinline[style=oxcaml]{handle_gen} handler, then we get an error:
\begin{lstlisting}[style=oxcaml]
  handle_await (fun ch ->
    handle_gen i (fun gh ->
      List.iter files (fun fl ->
        let s = await ch (File.read fl) in
        let i = int_of_string s in
        gen gh i)))
  (*@ \textcolor{red}{Error: let s = await ch (File.read fl) in} @*)
                       (*@ \textcolor{red}{\textasciicircum\textasciicircum} @*)
  (*@ \textcolor{red}{ This value is local and would escape its region.} @*)
\end{lstlisting}

@@

  module Project(L : Op) (R : Op) : sig
    val outl : Handler(Sum(L)(R)).t -> Handler(L).t

    val outr : Handler(Sum(L)(R)).t -> Handler(R).t
  end
\end{lstlisting}

@@

%\begin{lstlisting}[style=oxcaml]
%  module Half_reflection (L : Op) (R : Op) : sig
%    module L_eff = Reflection(L)
%    module R_eff = Reflection(R)
%    module Sum_eff = Reflection(Sum(L)(R))
%    module Proj = Project(L)(R)
%
%    let reify (f : Handler(L).t @ local -> Handler(R).t @ local -> 'a) lh =
%      let rec loop lh = function
%        | Term(Sum(L)(R)).Return ->
%            Half_term(L)(R).Return v
%        | Term(Sum(L)(R)).Op(Left op, k) ->
%            loop lh (k (L_eff.perform op lh))
%        | Term(Sum(L)(R)).Op(Right op, k) ->
%            Half_term(L)(R).Op(op, (fun res lh -> loop lh (k res)))
%      in
%      loop (Sum_eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))) lh
%
%    let rec reflect (f : Handler(L).t @ local -> 'a Half_term(L)(R).t) lh rh =
%      match f lh with
%      | Return x -> x
%      | Op(op, k) -> reflect (k (R_eff.perform op rh)) lh rh
%end
%\end{lstlisting}

Now we can use this machinery to conveniently build a handler of
\lstinline[style=oxcaml]{Gen} that also forwards operations from
\lstinline[style=oxcaml]{Await}:
\begin{lstlisting}[style=oxcaml]
  module Await_gen_eff = Half_reflection(Await)(Gen)

@@

  let handle_gen_in_await f ah =
      let rec loop = function
      | Half_term(Await)(Gen).Return () -> Finished
      | Half_term(Await)(Gen).Op(Gen i, k) -> More(i, fun ah -> loop (k () ah))
    in
    loop (Await_gen_eff.reify f ah)
\end{lstlisting}
where \lstinline[style=oxcaml]{handle_gen_in_await} has the type:
\begin{lstlisting}[style=oxcaml]
  val handle_gen_in_await :
    (Handler(Await).t @ local -> Handler(Gen).t @ local -> 'a)
    -> Handler(Await).t @ local
    -> aints
\end{lstlisting}

@@

  module Term1 (O : Op1) : sig
    type ('a, 'e) t =
      | Return : 'a @@ global -> 'a t
      | Op : ('r, 'e) O.t @@ global * ('r -> 'a t) -> 'a t
  end

  module Handler1(O : Op1) : sig
    type 'e t
  end
\end{lstlisting}
with a corresponding parameterized version of
\lstinline[style=oxcaml]{Reflection}. We can then define a parameterized
state effect:
\begin{lstlisting}[style=oxcaml]
  module State = struct
    type ('a, 'b) t =
      | Get : 'b t
      | Set : 'b -> unit t
  end
\end{lstlisting}
and write a handler that provides a parameterized interface for the
state effect:
\begin{lstlisting}[style=oxcaml]
  type 's handler

  val handle : ('s handler @ local -> 'a) @ local -> 's -> 'a
  val get : 's handler @ local -> 's
  val set : 's handler @ local -> 's -> unit
\end{lstlisting}

@@

We can implement the interface from Section \ref{sec:interface} directly
in terms of OCaml's untracked effect
handlers~\cite{Sivaramakrishnan2021retrofitting}. The trick is to
generate a fresh effect constructor for each reification, pass that
constructor around as part of the \lstinline[style=oxcaml]{handler}, and
then use it to perform operations during reflection. The full implementation
is given in Fig.\ref{fig:implementation}.

@@

module Reflection (O : Op) = struct

  let reify f =
    let module Handler =
      struct type 'a Effect.t += H : 'a O.t -> 'a Effect.t end
    in
    let handler : Handler(O).t =
      { perform = fun o -> Effect.perform (Handler.H o) }
    in
    match f handler with
    | v -> Return v
    | effect Handler.H o, k -> Op(o, fun x -> Effect.continue k x)

  let reify_local f = exclave
    let module Handler =
      struct type 'a Effect.t += H : 'a O.t -> 'a Effect.t end
    in
    let handler = { perform = fun o -> Effect.perform (Handler.H o) } in
    match f handler with
    | v -> Return v
    | effect Handler.H o, k -> Op(o, fun x -> Effect.continue k x)

@@

module Project (L : Op) (R : Op) = struct

  let outl {perform} =
    {perform = fun op -> perform (Sum(L)(R).Left op)}

  let outr {perform} =
    {perform = fun op -> perform (Sum(L)(R).Right op)}

@@

\pagebreak
\section{Related work}
\label{sec:related-work}

>>>>

  module State_eff = Reflection(State)
  open Term(State)

  let handle_state (f : Handler(State).t @ local -> 'a) (i : int) : 'a =
    let rec loop (s : int) : 'a Term(State).t @ local -> 'a  = function
      | Return x -> x
      | Op(Get, k) -> loop s (k s)
      | Op(Set s', k) -> loop s' (k ())
    in
    loop i (State_eff.reify_local f)
\end{lstlisting}
which calls \lstinline[style=oxcaml]{f} and passes it a local value of
type \lstinline[style=oxcaml]{Handler(State).t}. For the duration of the
call to \lstinline[style=oxcaml]{f} the ambient monad is extended with
two additional effectful operations, which can be used via reflection:
\begin{lstlisting}[style=oxcaml]
  let get (h : Handler(State).t @ local) : int =
    State_eff.perform Get h

@@

Locality tracks whether a value can 'escape' the current
\emph{region}. By default, the current region ends when a function
returns, so local values cannot be returned from functions.
\begin{lstlisting}[style=oxcaml]
  let leak_local x =
    let (o @ local) = Some x in
    o
  (*@ \textcolor{red}{Error: o} @*)
         (*@ \textcolor{red}{\textasciicircum} @*)
  (*@ \textcolor{red}{This value escapes its region.} @*)
\end{lstlisting}

@@

As shown in the introduction we can implement an effect handler for
state using this API. We start by defining the operations of the effect:
\begin{lstlisting}[style=oxcaml]
  module State = struct
    type 'a t =
      | Get : int t
      | Set : int -> unit t
  end
\end{lstlisting}
Then we instantiate the reflection isomorphism for that effect and use
it to write a handler:
\begin{lstlisting}[style=oxcaml]
  module State_eff = Reflection(State)
  open Term(State)

  let handle_state i f =
    let rec loop (s : int) = function
      | Return x -> x
      | Op(Get, k) -> loop s (k s)
      | Op(Set s', k) -> loop s' (k ())
    in
    loop i (State_eff.reify_local f)
\end{lstlisting}
Finally we add an alias for the handler type and some convenient
wrappers for the generic effects:
\begin{lstlisting}[style=oxcaml]
  type handler = Handler(State).t

@@

  val handle_state : int -> (handler @ local -> 'a) @ local -> 'a
  val get : handler @ local -> int
  val set : handler @ local -> int -> unit
\end{lstlisting}
Note that we are able to use \lstinline[style=oxcaml]{reify_local} to allow
the function passed to \lstinline[style=oxcaml]{handle_state} to be local.

@@

  let handle_state i f = f (ref i)
  let get r = !r
  let set r i = r := i
\end{lstlisting}
This implementation gives the same behaviour as the one built using
effect reflection. Crucially it is just as easy to reason about, which
in turn means it is just as easy to reason about as the state monad. In
some sense, the improved reasoning ability of the state monad over
ordinary mutable state is precisely captured by the notion of locality.

@@

The following function uses state to calculate the total of a list of
integers:
\begin{lstlisting}[style=oxcaml]
  let total l =
    handle_state 0 (fun h ->
      List.iter (fun x -> set h (get h + x)) l;
      get h)
\end{lstlisting}
This relies on the locality of the function parameter of
\lstinline[style=oxcaml]{List.iter}:
\begin{lstlisting}[style=oxcaml]
  val iter : ('a -> unit) @ local -> 'a list -> unit
\end{lstlisting}
The \lstinline[style=oxcaml]{local} annotation on the parameter ensures
that any function passed will not escape the current region. This allows
us to pass in a closure that closes over the
\lstinline[style=oxcaml]{local} handler \lstinline[style=oxcaml]{h} and
use it to perform some effects.

@@

A more realistic use of effect reflection is a \emph{generator}
effect. This effect has a single operation:
\begin{lstlisting}[style=oxcaml]
  module Gen = struct
    type 'a t =
      | Gen : int -> unit t
  end
\end{lstlisting}
Along with a handler that turns computations using
\lstinline[style=oxcaml]{Gen} into streams of integers.
\begin{lstlisting}[style=oxcaml]
  module Gen_eff = Reflection(Gen)
  open Term(Gen)

@@

  let handle_gen f =
    let rec loop = function
      | Return () -> Finished
      | Op(Gen i, k) -> More(i, fun () -> loop (k ()))
    in
    loop (Gen_eff.reify f)

\end{lstlisting}

@@

\begin{lstlisting}[style=oxcaml]
  module Await_eff = Reflection(Await)
  open Term(Await)

  let handle_await f =
    let rec loop = function
      | Return x -> Deferred.return x
      | Op(Await d, k) ->
          Deferred.bind d (fun x -> loop (k x))
    in
    loop (Await_eff.reify f)

@@

A key feature of algebraic effects is how easy they are to compose. We
can use state whilst using asynchronous I/O by simply nesting their
handlers:
\begin{lstlisting}[style=oxcaml]
  handle_await (fun ch ->
    handle_state i (fun sh ->
      List.iter (fun fl ->
        let s = await ch (File.read fl) in
        let i = int_of_string s in
        set sh (i + get sh)) files;
      get sh))
\end{lstlisting}

@@

If we tried to do something similar using our
\lstinline[style=oxcaml]{handle_gen} handler, then we get an error:
\begin{lstlisting}[style=oxcaml]
  handle_await (fun ch ->
    handle_gen (fun gh ->
      List.iter (fun fl ->
        let s = await ch (File.read fl) in
        let i = int_of_string s in
        gen gh i) files))
  (*@ \textcolor{red}{Error: let s = await ch (File.read fl) in} @*)
                       (*@ \textcolor{red}{\textasciicircum\textasciicircum} @*)
  (*@ \textcolor{red}{ This value is local and would escape its region.} @*)
\end{lstlisting}

@@

  module Project(L : Op) (R : Op) : sig
    val outl : Handler(Sum(L)(R)).t @ local -> Handler(L).t @ local

    val outr : Handler(Sum(L)(R)).t @ local -> Handler(R).t @ local
  end
\end{lstlisting}

@@

Now we can use this machinery to conveniently build a handler of
\lstinline[style=oxcaml]{Gen} that also forwards operations from
\lstinline[style=oxcaml]{Await}:
\begin{lstlisting}[style=oxcaml]
  module Await_gen_eff = Half_reflection(Await)(Gen)
  open Half_term(Await)(Gen)

@@

  let handle_gen_in_await f ah =
      let rec loop = function
      | Return () -> Finished
      | Op(Gen i, k) -> More(i, fun ah -> loop (k () ah))
    in
    loop (Await_gen_eff.reify f ah)
\end{lstlisting}
where \lstinline[style=oxcaml]{handle_gen_in_await} has the type:
\begin{lstlisting}[style=oxcaml]
  val handle_gen_in_await :
    (Handler(Await).t @ local -> Handler(Gen).t @ local -> unit)
    -> Handler(Await).t @ local
    -> aints
\end{lstlisting}

@@

  module Term1 (O : Op1) : sig
    type ('a, 'e) t =
      | Return : 'a @@ global -> ('a, 'e) t
      | Op : ('r, 'e) O.t @@ global * ('r -> ('a, 'e) t) -> ('a, 'e) t
  end

  module Handler1(O : Op1) : sig
    type 'e t
  end
\end{lstlisting}
with a corresponding parameterized version of
\lstinline[style=oxcaml]{Reflection}. We can then define a parameterized
state effect:
\begin{lstlisting}[style=oxcaml]
  module State = struct
    type ('a, 's) t =
      | Get : ('s, 's) t
      | Set : 's -> (unit, 's) t
  end
\end{lstlisting}
and write a handler that provides a parameterized interface for the
state effect:
\begin{lstlisting}[style=oxcaml]
  type 's handler

  val handle_state : 's -> ('s handler @ local -> 'a) @ local -> 'a
  val get : 's handler @ local -> 's
  val set : 's handler @ local -> 's -> unit
\end{lstlisting}

@@

We can implement the interface from Section \ref{sec:interface} directly
in terms of OCaml's untracked effect
handlers~\cite{Sivaramakrishnan2021retrofitting}\footnote{For
  convenience, we use the new handler syntax from OCaml 5.3. However,
  the most recently released version of OxCaml is based on OCaml 5.2 so
  the actual implementation uses
  \lstinline[style=oxcaml]{Effect.Deep.match_with} instead}. The trick
is to generate a fresh effect constructor for each reification, pass
that constructor around as part of the
\lstinline[style=oxcaml]{handler}, and then use it to perform operations
during reflection. The full implementation is given in
Fig.\ref{fig:implementation}.

@@

module Reflection (O : Op) = struct
  open Term(O)
  open Handler(O)

  let reify f =
    let module Eff =
      struct type 'a Effect.t += C : 'a O.t -> 'a Effect.t end
    in
    let handler =
      { perform = fun o -> Effect.perform (Eff.C o) }
    in
    match f handler with
    | v -> Return v
    | effect Eff.C o, k -> Op(o, fun x -> Effect.continue k x)

  let reify_local f = exclave
    let module Eff =
      struct type 'a Effect.t += C : 'a O.t -> 'a Effect.t end
    in
    let handler = { perform = fun o -> Effect.perform (Eff.C o) } in
    match f handler with
    | v -> Return v
    | effect Eff.C o, k -> Op(o, fun x -> Effect.continue k x)

@@

module Project (L : Op) (R : Op) = struct
  open Sum(L)(R)

  let outl ({perform} : Handler(Sum(L)(R)).t) : Handler(L).t =
    exclave {perform = fun op -> perform (Left op)}

  let outr ({perform} : Handler(Sum(L)(R)).t) : Handler(R).t =
    exclave {perform = fun op -> perform (Right op)}

@@

\section{Related work}
\label{sec:related-work}
